---
title: "Annotations Example"
author: "Moriah"
date: "9/26/2019"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include = FALSE, warnings = FALSE, echo = TRUE}
##install packages##
library(dplyr)    #data manipulation
library(datasets)    #collection of datasets
library(rmarkdown)    #markdown
```

The dataframe **sleep** from the package **datasets** contains 20 observations on 3 variables and represent the results of an experiment in which two groups of students were given different treatments and the change in the hours of sleep that they got was recorded.   
    * **extra**: change in hours of sleep (type = numeric)  
    * **group**: drug given (type = factor)  
    * **ID**: patient ID (type = factor)
```{r, echo = TRUE}
head(sleep)  #preview data

#change column name
#in order to prevent confusion when using
#group_by() in piping
colnames(sleep)[colnames(sleep)=="group"] <- "treatment"    #rename column
head(sleep)    #see that column name is changed
```
Let's take a look at the average effects of each treatment.
```{r, echo = TRUE}
##find mean change in sleep for each treatment##
averages <- sleep %>%    #store results to object named 'averages'
  group_by(treatment) %>%    #group observations by treatment
  summarise(avg = mean(extra))    #calculate mean change in hours of sleep
averages    #see results
```

\pagebreak

Now let's compare the boxplots of the effects of each treatment.
```{r, echo = TRUE}
##create a boxplot##
boxplot(extra ~ treatment, data = sleep,
        xlab = "Treatment",    #label x-axis
        ylab = "Change in Sleep (Hrs)",    #label y-axis
        main = "Treatment Effects",    #title boxplots
        names = c("1", "2"),    #label treatment types
        col = c("skyblue", "plum"),    #choose color for each boxplots
        ylim = c(-2,6)    #increase range of y-axis
        )
```

\pagebreak

Finally, let's create a simple function called **eval_effect** that returns whether a value represents an increase, decrease, or no change. Then, let's apply that function to the column 'extra' in the sleep dataset and create a new column called 'effect'.
```{r, echo = TRUE}
##create function eval_effect##
#input: value representing observed change
#output: statement about type of change
eval_effect <- function(observation){
  if (observation > 0){    #label a positive value as an increase
    return("increase")}
  else if (observation < 0){    #label a negative value as a decrease
    return("decrease")}
  else {                    #label a zero value as no change
    return("no change")}
}

##effect column##
#create column
sleep$effect <- apply(sleep[1], 1, eval_effect)    #apply eval_effect function to observations
#view data with new column
sleep
```